---
title: "Homelessness in Los Angeles"
output: 
  html_notebook: 
    theme: flatly
    code_folding: hide
---

```{r setup, include=F}

knitr::opts_chunk$set(warning=FALSE, message=FALSE)

library(extrafont)
extrafont::loadfonts(device="win")
library(tidyverse)
library(lubridate)



```

**“Homelessness kills.”** These words are written on bench advertisements found across Los Angeles which coincidentally often serve as a place to sleep for some of the LA homeless population. The trade off between making a powerful statement and reminding those experiencing homelessness of their raised mortality... While the…[insensitivity of ads vs importance of issue]. I witness this first-hand every day on my commute through Venice and Santa Monica.  Just a few streets from the front door of my house (almost equi-distant from Sony Pictures Studios) there is an ever growing shanty town beneath the freeway underpass.  In Los Angeles, stepping over fellow Angelenos on the sidewalk has become as banal an experience as rain in Ireland.  Once every few months, county officials clear out the underpass in the morning.  Hours later the shopping trolleys, battered bicycles, and the tents return like the tide.  It isn’t clear whether authorities do this as an act of goodwill to our homeless residents – giving their space a much needed cleaning – or yet another prodding in the hopes that they will move somewhere else.  
As the pervasiveness of homelessness in Los Angeles became painfully evident, I grew curious about what we as a society, and more importantly what policy-makers know about homelessness.  How many people are homeless?  What is causing people to end up on the streets?  How do we help people end their bout of homelessness?  What programs are effective at returning people to homes? In addition to this, defining who is “homeless” becomes somewhat difficult when you get into the weeds e.g. how long does one have to spend without a home to count? A night? A week?  

## How many people are homeless?

It may be a sign of the difficulty of this problem that this simple question is very hard to answer.  If you have seen any news articles about the homelessness crisis, you may have seen numbers around 50,000.  The trouble is  – how can we accurately count a population which changes day-by-day both geographically (e.g. people traveling from Santa Monica to Skid Row)  and definitionally (e.g. people finding a place to stay temporarily)?  The Los Angeles Homeless Services Authority (LAHSA) [report there were 58,936 people experiencing homelessness](https://www.lahsa.org/documents?id=3437-2019-greater-los-angeles-homeless-count-presentation.pdf) in Los Angeles County at the Point-In-Time (PIT) Homelessness count in 2019.  This is an increase of 12% from the homeless count in 2018.
Unfortunately the PIT data comes with a few caveats.  While it is a valuable resource, it only gives us an estimate of how many people experience homelessness on a given night, which is a very narrow snapshot in time.  In 2018, researchers from the Economic Roundtable estimated that [the annual population experiencing homelessness was just over 100,000 people](https://economicrt.org/publication/estimating-the-annual-size-of-the-homeless-population/) – this was nearly double the PIT estimate in the same year. On top of this, [the results of the homeless count are not accurate enough to identify year-to-year changes](https://economicrt.org/publication/who-counts/), which means we cannot be confident about the 12% increase mentioned above. This is problematic because policy responses depend on an accurate understanding of the situation to make an informed decision to try to address the problem.

## What is causing people to end up on the streets?

### Precarious Housing

```{r poverty-status, fig.width = 14, fig.height = 8}

# function to increase vertical spacing between legend keys
# @clauswilke: https://stackoverflow.com/questions/11366964/is-there-a-way-to-change-the-spacing-between-legend-items-in-ggplot2
draw_key_polygon3 <- function(data, params, size) {
  lwd <- min(data$size, min(size) / 4)

  grid::rectGrob(
    width = grid::unit(0.6, "npc"),
    height = grid::unit(0.6, "npc"),
    gp = grid::gpar(
      col = data$colour,
      fill = alpha(data$fill, data$alpha),
      lty = data$linetype,
      lwd = lwd * .pt,
      linejoin = "mitre"
    ))
}

# register new key drawing function, 
# the effect is global & persistent throughout the R session
GeomBar$draw_key = draw_key_polygon3

# Population around the poverty line
census_files = list.files("data/PovertyStatus", pattern = "[0-9]\\.csv$", full.names = T)

map_dfr(census_files, function(dataset){
  
  # Identify related metadata file
  metadata = str_replace(dataset, "\\.csv$", "_metadata.csv")
  # Identify file year
  census_year = str_extract(dataset, "[0-9]{4}(?=\\.csv$)")
  
  # 1. Select the vars we are interested in
  poverty_vars = read_csv(metadata) %>% 
    # Remove variables which do not pertain to total population poverty level groupings
    filter(!str_detect(id, "Margin of Error"), !str_detect(id, "!!(SEX|AGE|RACE|LIVING|EDUCATION|NATIVITY|DISABILITY|WORK|CITIZENSHIP)|Hispanic")) %>% 
    # Remove empty column
    select(-X3) %>% 
    # Remove extraneous text in the id var
    mutate(id = str_remove_all(id, "Estimate!!"),
           id = str_remove_all(id, "!!Population for whom poverty status is determined"),
           id = str_replace_all(id, "\\spercent ", "%\n"))
  
  # 2. Filter census data on these vars
  poverty_estimates = read_csv(dataset) %>% 
    # Remove first row (variable labels)
    slice(-1) %>% 
    # Select vars from above table
    select(poverty_vars$GEO_ID) %>% 
    rename_at(2, ~paste("population")) %>% 
    # Gather vars except for grouping and location vars
    gather(key = "group", value = "estimate", -1:-2) %>% 
    # Transformation
    # - Convert estimate to numeric percentage
    # - Calculate estimate in nominal terms
    # - Replace poverty level grouping ids with explanatory labels
    # - Order poverty level grouping variable from smallest to largest
    # - Add year variable
    mutate(estimate_pct = as.numeric(estimate)/100,
           estimate_n = as.numeric(population)*estimate_pct,
           group = str_replace_all(group, 
                                   pattern = filter(poverty_vars, str_detect(id, "poverty")) %>% pull(GEO_ID), 
                                   replacement = filter(poverty_vars, str_detect(id, "poverty")) %>% pull(id)),
           group = factor(group, levels = c("Less than 50%\nof the poverty level", 
                                            "Less than 100%\nof the poverty level", 
                                            "Less than 125%\nof the poverty level"), ordered = T),
           group = fct_rev(group),
           year = as.numeric(census_year),
           estimate_n = estimate_n - lag(estimate_n, default = 0))
}) %>% 
  # Plot
  ggplot(., aes(x = year, y = estimate_n, fill = group))+
  geom_col(position = "stack")+
  scale_x_continuous(breaks = 2010:2018)+
  scale_y_continuous(labels = scales::comma)+
  scale_fill_manual(values = c("#ffa600", "#ff7715", "#ff3b30"), name = "")+
  labs(y = "", x = "", 
       title = "Millions experience poverty in Los Angeles County each year",
       # subtitle = "Based on US Census specified federal poverty levels",
       subtitle = bquote('Based on US Census specified' ~ bold('federal') ~ 'poverty levels'),
       caption = "Source: American Community Survey, TableID S1703")+
  # facet_wrap(~year, nrow = 1, strip.position = "bottom")+
  theme_minimal()+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        text = element_text(family = "Roboto Medium", size = 14),
        plot.title = element_text(face = "bold", size = 18), 
        plot.subtitle = element_text(size = 14),
        legend.key.size = unit(1.25, "cm"))

#ff3b30
#ff7715
#ffa600

```

So we have millions of people in Los Angeles County experiencing poverty each year. According to a [homelessness meta-analysis by the Economic Roundtable](https://livingwage.mit.edu/counties/06037), **half of families in poverty spend 90% of their income or more on housing**. That leaves *very little* slack for sudden shortages in income

And this is likely not even the full story.  The Census uses Federal Poverty Guidelines to determine the number of people in these various income groups and that does not take into account the local cost of living.  But how much money do people need to live in Los Angeles?  [Researchers at MIT have calculated what is called a "Living wage"](https://livingwage.mit.edu/counties/06037) across the country which takes into account how far the same dollar goes across the country all the way down to the county level.  

```{r fpg-vs-living-wage, fig.width = 14, fig.height = 8}
## What is the poverty level?
poverty_guidelines = tibble(
  household_mem = 1:8,
  poverty_lev = c(12490, 16910, 21330, 25750, 30170, 34590, 39010, 43430),
  living_wage = c(25680,39844,49118,53871,62444, 97000, 113000, 130000)
)

poverty_guidelines %>% 
  ggplot(., aes(x = household_mem, y = poverty_lev))+
  geom_line(colour = "#ff3b30")+
  geom_line(aes(y = poverty_lev*1.25), colour = "#ff3b30", linetype = "dashed")+
  geom_line(aes(y = living_wage), colour = "mediumseagreen")+
  # geom_hline(yintercept = annual_studio_rent, linetype = "dashed")+
  scale_x_continuous(breaks = 1:8)+
  scale_y_continuous(limits = c(0, 65000), breaks = c(0, 15000, 30000, 45000, 60000), labels = scales::dollar)+
  labs(x = "Persons in Family/Household", y = "Income\n",
       title = "Federal Poverty Guidelines do not come close to the minimum cost of living in Los Angeles County",
       subtitle = "Federal Poverty Guidelines (FPG) vs. Living Wage income, by household size",
       caption = "Sources: US Dept. Health & Human Services\nMIT Living Wage Calculator\n\n*Required income varies by household composition.  Using minimum income required by household size")+
  theme_minimal()+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        text = element_text(family = "Roboto Medium", size = 14),
        plot.title = element_text(face = "bold", size = 18), 
        plot.subtitle = element_text(size = 14),
        plot.caption = element_text(family = "Roboto", size = 12),
        legend.key.size = unit(1.25, "cm"))+
  annotate("text", x = 8.275, y = 43600, label = "100% FPG")+
  annotate("text", x = 8.275, y = 54500, label = "125% FPG")+
  annotate("text", x = 6.1, y = 62600, label = "Income required to meet basic household needs*")


```


### Criminalizing Homelessness

That there are people who do not have somewhere to live, or even somewhere to sleep is tragic.  What is even worse is the criminalization of their circumstances.   

```{r}
datadir = "C:/Users/Mark/Documents/DataScience/Projects/Homelessness/data/LA Crime"
dataURL = "https://data.lacity.org/api/views/yru6-6re4/rows.csv?accessType=DOWNLOAD&bom=true&format=true"
arrdata = read_csv(file.path(datadir, "Arrest_Data_from_2010_to_Present.csv")) %>%
# arrdata = read_csv(dataURL) %>%
  mutate(`Arrest Date` = mdy(`Arrest Date`))

arrdata_homeless = arrdata %>%
  filter(str_detect(Charge, "41\\.*18")) %>%
  mutate(crime = case_when(str_detect(Charge, "41\\.*18")~"Loitering"),
         year = year(`Arrest Date`),
         month_year = floor_date(`Arrest Date`, unit = "month"),
         lat = as.numeric(str_extract(Location, "[^\\(](.+?)(?=,)")),
         lon = as.numeric(str_extract(Location, "(?<=, ).*[^)]")),
         descent = str_replace_all(`Descent Code`, c("O" = "Other", "A" = "Other Asian", "B" = "Black", "C" = "Chinese",
                                                     "D" = "Cambodian", "F" = "Filipino", "G" = "Guamanian",
                                                     "H" = "Hispanic", "I" = "Native American", "J" = "Japanese",
                                                     "K" = "Korean", "L" = "Laotian", "P" = "Pacific Islander",
                                                     "S" = "Samoan", "U" = "Hawaiian", "V" = "Vietnamese", "W" = "White",
                                                     "X" = "Unknown", "Z" = "Indian")))

arrdata_homeless %>% 
  count(month_year, crime, name = "arrests") %>% 
  ggplot(aes(x = month_year, y = arrests))+
  geom_col()+
  ggtitle("Loitering in Los Angeles 2010-2018", subtitle = "Monthly arrests for loitering charges (L.A.M.C. § 48.18)")+
  labs(x = "", y = "Monthly Arrests",
       caption = "Source: data.lacity.org")+
  theme_minimal()+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        text = element_text(family = "Roboto Medium", size = 14),
        plot.title = element_text(face = "bold", size = 18), 
        plot.subtitle = element_text(size = 14),
        plot.caption = element_text(family = "Roboto", size = 12),
        legend.key.size = unit(1.25, "cm"))
```


